Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-12 manuscript/climate_sensitivity_draft.Rmd
update table 1 and text to remove DBH test
#99

- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA) to test hypotheses on how tree size, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

- Individual-level drought resistance decreased with tree height, which was the dominant size-related variable affecting drought response. Resistance was greater  among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon desiccation. However, there was substantial variation in the best predictor variables across the three drought periods.

@@

Here, we examine how tree size, microenvironment characteristics, and species' traits collectively shape drought responses. We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA. First, we focus on the role of tree size and its interaction with microenvironment. We test whether, consistent with most forests globally, larger-diameter trees tend to have lower drought resistance ($Rt$) in this forest, which is in a region (eastern North America) represented by only two studies in the global review of @bennett_larger_2015. 
We then test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.

@@

Tree size, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of larger trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height rather than crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.  The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

>>>>

- We analyzed tree-ring records for twelve species that comprise 97% of the woody productivity of the 25.6-ha ForestGEO plot in an oak-hickory forest of northern Virginia (USA) to test hypotheses on how tree height, microenvironment characteristics, and species' traits shaped drought responses across the three strongest regional droughts over a 60-year period (1950 - 2009).

- Individual-level drought resistance decreased with tree height. Resistance was greater among species whose leaves lost turgor (wilted) at more negative water potentials, and whose leaves experienced less shrinkage upon desiccation. However, there was substantial variation in the best predictor variables across the three drought periods.

@@

Here, we examine how tree height, microenvironment characteristics, and species' traits collectively shape drought responses. We test a series of hypotheses and associated specific predictions (Table 1) based on the combination of tree-ring records from three droughts (1966, 1977, 1999), species functional and hydraulic trait measurements, and census data from a large forest dynamics plot in Virginia, USA. First, we focus on the role of tree height and its interaction with microenvironment. 
We test hypotheses designed to disentangle the relative importance of tree height; crown exposure; and soil water availability, which should be greater for larger trees in dry but not in perpetually wet microsites. 
Second, we focus on the role of species' functional and hydraulic traits, testing the hypothesis that species’ traits-–particularly leaf hydraulic traits-–predict $Rt$. We test predictions that drought resistance is higher in ring-porous than semi-ring and diffuse-porous species, that it is correlated with wood density--either postively [@greenwood_tree_2017] or negatively [@hoffmann_hydraulic_2011]-- and positively correlated with $LMA$, and that hydraulic leaf traits including $PLA_{dry}$ and $\pi_{tlp}$ are better predictors.

@@

Tree height, microenvironment, and hydraulic traits shaped tree growth responses across three droughts at our study site (Table 1). The greater susceptibility of taller trees to drought, similar to forests worldwide [@bennett_larger_2015], was driven primarily by their height with modest additional affect of crown exposure [@liu_effect_1993; @stovall_tree_2019]. We found only a marginal additional effect of crown exposure, with a tendency for lowest $Rt$ among the most exposed (dominant) and suppressed trees.  The negative effect of height on $Rt$ held after accounting for species' traits. There was no evidence that soil water availability increased drought resistance; in contrast, trees in wetter topographic positions had lower $Rt$ [@zuleta_drought-induced_2017; @stovall_tree_2019], and the larger potential rooting volume of large trees provided no advantage in the drier microenvironments. Drought resistance was not consistently linked to species' $LMA$, wood density, or xylem type (ring- vs. diffuse porous), but was negatively correlated with leaf hydraulic traits ($PLA_{dry}$, $\pi_{tlp}$) in the top overall model and the top models for two of the three individual droughts. This is the first report to our knowledge linking $PLA_{dry}$ and $\pi_{tlp}$ to growth reduction during drought. The direction of responses was mostly consistent across droughts, supporting the premise that they were driven by fundamental physiological mechanisms. However, the strengths of each predictor varied across droughts (Tables 4-5), indicating that drought characteristics interact with tree size, microenvironment, and traits to shape which individuals are most affected. These findings advance our knowledge of the factors that make trees vulnerable to growth declines during drought--and, by extension, likely make them more vulnerable to mortality [@sapes_plant_2019].

### 2020-07-12 manuscript/tables&figures.Rmd
update table 1 and text to remove DBH test
#99

\blandscape
```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Larger, taller trees have lower Rt.", 1, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Trees with more exposed crowns have lower Rt.", 3, 4, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 5, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 6, 10, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1.2cm") %>%
  column_spec(6, width = "2.4cm") 
 # kableExtra::landscape()
```

>>>>

\blandscape
```{r Table 1, eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("tables_figures/publication/table1_hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Table 1. Summary of hypotheses, corresponding specific predictions, and results.") %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 8.5, protect_latex = TRUE) %>%
  pack_rows("Tree size and microenvironment", 1, 5, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Taller trees have lower Rt.", 1, 1, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Trees with more exposed crowns have lower Rt.", 2, 2, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Small trees (lower root volume) in drier microhabitats have lower Rt.", 3, 3, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  pack_rows("Species traits", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE) %>%
  pack_rows("Species' traits--particularly leaf hydraulic traits--predict Rt.", 4, 8, latex_gap_space = "1em", colnum = 1, hline_before = FALSE, bold = FALSE, italic = TRUE) %>%
  add_header_above(c(" ", "Prediction supported?" = 4, " ")) %>%
  column_spec(1, width = "12cm") %>%
  column_spec(2:5, width = "1.2cm") %>%
  column_spec(6, width = "2.4cm") 
 # kableExtra::landscape()
```
